# Three ways to quantify child development {#sec:threeways}

This section distinguishes three different principles that are being
used to quantify child development: 

* Age-based measurement (\@ref(sec:agebased))
* Score-based measurement (\@ref(sec:scorebased))
* Unit-based measurement (\@ref(sec:unitbased))

## Age-based measurement of development {#sec:agebased}

### Motivation for age-based measurement

The motivation for measuring child development is to be able to
describe the behavior expected at a selected age. @gesell1943 (p. 89)
formulates this goal as follows:

> We think of behavior in terms of age, and we think of age in terms
> of behavior. For any selected age it is possible to sketch a portrait
> which delineates the behavior characteristics typical of the age.

There is a large literature that quantifies development in terms of
ages at which the child is expected to show a specific behavior. For
example, in Shirley's motor data in Table \@ref(tab:shirley), we may
summarize the ages at which each of the 21 children enters a given
stage.

```{r agebased, echo = FALSE, fig.height = 3, fig.cap = '(ref:agebased)', warning = FALSE}
load(file = file.path("data/shirley.rda"))

d <- shirley %>%
  mutate(name = factor(name, as.character(name))) %>%
  gather(key = "stage", value = "age", -name, -sex) %>%
  mutate(
    stage = factor(stage, unique(stage)),
    score = rep(1:4, each = 21))
ggplot(d, aes(age, stage, group = name, colour = name)) +
  geom_point(cex = 3, pch = 1) + 
  geom_point() + 
  theme_light() + 
  guides(colour = FALSE) + 
  labs(x = "Age (weeks)", y = "Stage")
```

(ref:agebased) Ages at which 21 children achieve four motor development milestones. 

### Age equivalent as difficulty {#sec:ageequivalent}

Since age and development are so intimately related, the *difficulty*
of a given developmental milestone is traditionally expressed in terms
of the age at which children achieve it. For example,
@stott1967 (p. 25) defines the *age equivalent* and its use for
measurement, as follows:

> The age equivalent of a particular stage is simply the average age 
> at which children reach that particular stage. 

```{r agebox, echo = FALSE, fig.height = 3, fig.cap = '(ref:agebox)', warning = FALSE}
means <- aggregate(age ~  stage, d, mean)
ggplot(d, aes(y = age, x = stage, group = stage, colour = name)) +
  geom_boxplot(col = "skyblue", fill = "transparent", width = 0.3) +
  geom_point(cex = 3, pch = 1) + 
  geom_point() + 
  stat_summary(fun.y = mean, colour = "black", geom = "point", 
               shape = 4, size = 3, show_legend = FALSE) +
  theme_light() + 
  guides(colour = FALSE) + 
  labs(y = "Age (weeks)", x = "Stage") +
  coord_flip()
```

(ref:agebox) Boxplot and mean (symbol `x`) of the ages at which 21 children achieve four motor development milestones.

Figure \@ref(fig:agebox) adds boxplot and the mean age at which the
children enter the four stages. The difficulty of the stages can thus
be expressed as age equivalents: 16.1 weeks for stepping, 27.2 weeks
for standing, 43.3 weeks for walking with help and 63.3 weeks for
walking alone. 

Alternatives to the mean age equivalent include the ages at which 10,
50 or 90 percent of the children achieve the milestone. This line of
reasoning has led to the widely used concept of *developmental age* as
a measure of a child's development (in body size or motor skill or
psychological function) expressed in terms of age norms.

### Limitations of age-based measurement 

Age-based measurement is easy to understand, but is not without 
pitfalls:

1. Age-based measurement can inform us whether a child is earlier or
later for a given milestone. However, it does not tell us what
behaviors are characteristic for a child of a given age.

2. Age-based measurement cannot exists without an age norm. When there
is no norms, we cannot quantify development. Age-based measurement
intertwines quantification of development with evaluation of
development.

3. Age-based measurement only works at the item level, and is 
cumbersome to apply with multiple items.

## Score-based measurement of development {#sec:scorebased}

### Motivation for score-based measurement

Score-based measurement takes the responses on multiple items and
counts the total number of passed items as a measure of child's
development. This approach takes multiple measurement into account,
hence leading to a more stable measurement. 

If items are ordered in difficulty, so one may skip items that are too
easy, and stop when items become too difficult for the child, for
example as age-forms. In such cases, we cannot simply interpret the
sum score of a measure of development, and need to correct for the
subset of items that was actually administered. The usual working
assumption is that the child would have passed all easier items and
failed on all more difficult items. This procedure is typically
repeated for different domains, e.g. motor, cognitive, and so on.

### Example of score-based measurement

```{r scoreplot, echo = FALSE, fig.asp = 0.6, fig.cap = '(ref:scoreplot)', warning = FALSE}
levels(d$stage) <- as.character(0:3)
ggplot(d, aes(age, stage, group = name, colour = name)) +
  geom_step() + 
  geom_point() + 
  theme_light() + 
  facet_wrap(~ name, nrow = 3) +
  guides(colour = FALSE) + 
  labs(x = "Age (weeks)", y = "Gross-motor score")
```

(ref:scoreplot) Same data as in Figure \@ref(fig:stepplot), but now with 
the vertical axis representing gross-motor score.

Figure \@ref(fig:scoreplot) is a gross-motor score calculated as the
number of items passed. It varies from 0 to 3. 

The plot suggests that the difference in development between scores 0
and 1 is the same as the difference in development between, say,
scores 2 and 3. *This is not true*. For example, suppose that we
express difficulty of the item as an age-equivalent. From section
\@ref(sec:ageequivalent) we see that the difference in difficulty
between stepping and standing is 27.2 - 16.1 = 11.1 weeks, whereas the
difference between walking alone and walking with help is 63.3 - 43.3
= 20 weeks. Thus, according to age equivalents scores 0 and 1 should be 
closer to each other, and scores 2 and 3 should be drawn wider apart.

### Limitations of score-based measurement

Score-based measurement is today's dominant approach, but is not
without conceptual and logistical issues.

1. The total score depends not only on the (true) developmental status
of the child, but also on the set of items actually administered. If
an item is skipped or added, the sum score cannot be interpreted
anymore as a measure of the child's developmental status.

2. It is not possible to compare scores made by different instruments.
Some instruments allows conversion to age-conditional scores. However,
the sample used to derive such conversions are typically restricted to
one instrument, and does not extend to measurements made by other
instruments.

3. Domains are hard to separate. For example, some cognitive items tap
into fine motor capabilities, and vice versa. There are different ways
to define domains, so domain interpretation varies by instrument.

4. Administration of a full test may take substantial time. The
materials are often proprietary and sometimes costly.

## Unit-based measurement of development {#sec:unitbased}

Unit-based measurement starts from the premise that the measurement
scale must have a constant unit. If that holds, then the same
difference between two values has the same interpretation anywhere on
the scale. Familiar examples of such scales includes temperature and
dates. For example, the difference between two dates corresponds to
the same amount of time, irrespective of whether this is now, in the
past or in the future. The interval scale has a fixed unit and an
arbitrary zero.

In the ideal situation, we would have a unit for child development.
Let us name this unit *D*, for development. Unit *D* should have the
following desirable properties:

- Difference between two *D*-scores measured on the same child at
different ages quantifies the amount of development during the
interval

- Difference between the *D*-score measured on two different children
of the same age quantifies the difference in development between them

- Comparing the *D*-score to an age-dependent reference identifies
abnormal development

The interpretation of the difference between two *D*-score is the same
across the entire scale, just like differences in height or weight.
Note that the unit would allow us to answer all questions raised in section @\ref(sec:questions).

The development of a unit for development is still in its infancy.
However, if we succeed in creating an interval scale for development,
then an enormous arsenal of technique developed for quantitative
variables opens up to measure, track and analyze child development.

## Role of unit-based measurement {#sec:whyunit}

This section reviewed age-based, score-based and unit-based approaches
to the measurement of child development. Age-based measurement
expresses development in age-equivalents, whose precise definition
depends on the reference population. Score-based measurement
quantifies development by summing the number of passes, but different
instruments make different selections of items, so the measurements
are unique to the instruments, and cannot be compared across
instruments. Neither approach provides a unit for child development.

In order to make progress, what we need is the definition of a unit for
child developent, and an operationalization how each measurement item
related to the unit. If we success in this, we are able to construct
different instruments that yields measurement in a common metric. The
next chapter will introduce the basic concepts for this new approach.
