# Three ways to quantify child development {#sec:threeways}

This section distinguishes three different principles that are being
used to quantify child development: 

* Age-based measurement (\@ref(sec:agebased))
* Score-based measurement (\@ref(sec:scorebased))
* Unit-based measurement (\@ref(sec:unitbased))

## Age-based measurement of development {#sec:agebased}

### Motivation for age-based measurement

The motivation for measuring child development is to be able to
describe the behavior expected at a selected age. @gesell1943 (p. 89)
formulates this goal as follows:

> We think of behavior in terms of age, and we think of age in terms
> of behavior. For any selected age it is possible to sketch a portrait
> which delineates the behavior characteristics typical of the age.

There is a large literature that quantifies development in terms of
ages at which the child is expected to show a specific behavior. For
example, in Shirley's motor data in Table \@ref(tab:shirley), we may
summarize the ages at which each of the 21 children enters a given
stage.

```{r agebased, echo = FALSE, fig.height = 3, fig.cap = '(ref:agebased)', warning = FALSE}
load(file = file.path("data/shirley.rda"))

d <- shirley %>%
  mutate(name = factor(name, as.character(name))) %>%
  gather(key = "stage", value = "age", -name, -sex) %>%
  mutate(
    stage = factor(stage, unique(stage)),
    score = rep(1:4, each = 21))
ggplot(d, aes(age, stage, group = name, colour = name)) +
  geom_point(cex = 3, pch = 1) + 
  geom_point() + 
  theme_light() + 
  guides(colour = FALSE) + 
  labs(x = "Age (weeks)", y = "Stage")
```

(ref:agebased) Ages at which 21 children achieve four motor development milestones. 

### Age equivalent as difficulty {#sec:ageequivalent}

Since age and development are so intimately related, the *difficulty*
of a given developmental milestone is traditionally expressed in terms
of the age at which children achieve it. For example,
@stott1967 (p. 25) defines the *age equivalent* and its use for
measurement, as follows:

> The age equivalent of a particular stage is simply the average age 
> at which children reach that particular stage. 

```{r agebox, echo = FALSE, fig.height = 3, fig.cap = '(ref:agebox)', warning = FALSE}
means <- aggregate(age ~  stage, d, mean)
ggplot(d, aes(y = age, x = stage, group = stage, colour = name)) +
  geom_boxplot(col = "skyblue", fill = "transparent", width = 0.3) +
  geom_point(cex = 3, pch = 1) + 
  geom_point() + 
  stat_summary(fun.y = mean, colour = "black", geom = "point", 
               shape = 4, size = 3, show_legend = FALSE) +
  theme_light() + 
  guides(colour = FALSE) + 
  labs(y = "Age (weeks)", x = "Stage") +
  coord_flip()
```

(ref:agebox) Boxplot and mean (symbol `x`) of the ages at which 21 children achieve four motor development milestones.

Figure \@ref(fig:agebox) adds boxplot and the mean age at which the
children enter the four stages. The difficulty of the stages can thus
be expressed as age equivalents: 16.1 weeks for stepping, 27.2 weeks
for standing, 43.3 weeks for walking with help and 63.3 weeks for
walking alone. 

Alternatives to the mean age equivalent include the ages at which 10,
50 or 90 percent of the children achieve the milestone. This line of
reasoning has led to the widely used concept of *developmental age* as
a measure of a child's development (in body size or motor skill or
psychological function) expressed in terms of age norms.

### Limitations of age-based measurement 

Age-based measurement is easy to understand, but is not without 
pitfalls:

1. Age-based measurement can inform us whether a child is earlier or
later for a given milestone. However, it does not tell us what
behaviors are characteristic for a child of a given age.

2. Age-based measurement cannot exists without an age norm. When there
is no norms, we cannot quantify development. Age-based measurement
intertwines quantification of development with evaluation of
development.

3. Age-based measurement only works at the item level, and is 
cumbersome to apply with multiple items.

## Score-based measurement of development {#sec:scorebased}

### Motivation for score-based measurement

Score-based measurement takes the responses on multiple items and
counts the total number of passed items as a measure of child's
development. This approach takes multiple measurement into account,
hence leading to a more stable measurement. 

If items are ordered in difficulty, so one may skip items that are too
easy, and stop when items become too difficult for the child, for
example as age-forms. In such cases, we cannot simply interpret the
sum score of a measure of development, and need to correct for the
subset of items that was actually administered. The usual working
assumption is that the child would have passed all easier items and
failed on all more difficult items. This procedure is typically
repeated for different domains, e.g. motor, cognitive, and so on.

### Example of score-based measurement

```{r scoreplot, echo = FALSE, fig.asp = 0.6, fig.cap = '(ref:scoreplot)', warning = FALSE}
levels(d$stage) <- as.character(0:3)
ggplot(d, aes(age, stage, group = name, colour = name)) +
  geom_step() + 
  geom_point() + 
  theme_light() + 
  facet_wrap(~ name, nrow = 3) +
  guides(colour = FALSE) + 
  labs(x = "Age (weeks)", y = "Gross-motor score")
```

(ref:scoreplot) Same data as in Figure \@ref(fig:stepplot), but now with 
the vertical axis representing gross-motor score.

Figure \@ref(fig:scoreplot) is a gross-motor score calculated as the
number of items passed. It varies from 0 to 3. 

The plot suggests that the difference in development between scores 0
and 1 is the same as the difference in development between, say,
scores 2 and 3. *This is not true*. For example, suppose that we
express difficulty of the item as an age-equivalent. From section
\@ref(sec:ageequivalent) we see that the difference in difficulty
between stepping and standing is 27.2 - 16.1 = 11.1 weeks, whereas the
difference between walking alone and walking with help is 63.3 - 43.3
= 20 weeks. Thus, according to age equivalents scores 0 and 1 should be 
closer to each other, and scores 2 and 3 should be drawn wider apart.

### Limitations of score-based measurement

Score-based measurement is today's dominant approach, but is not
without conceptual and logistical issues.

1. The total score depends not only on the (true) developmental status
of the child, but also on the set of items actually administered. If
an item is skipped or added, the sum score cannot be interpreted
anymore as a measure of the child's developmental status.

2. It is not possible to compare scores made by different instruments.
Some instruments allows conversion to age-conditional scores. However,
the sample used to derive such conversions are typically restricted to
one instrument, and does not extend to measurements made by other
instruments.

3. Domains are hard to separate. For example, some cognitive items tap
into fine motor capabilities, and vice versa. There are different ways
to define domains, so domain interpretation varies by instrument.

4. Administration of a full test may take substantial time. The
materials are often proprietary and sometimes costly.

## Unit-based measurement of development {#sec:unitbased}

### Motivation for unit-based measurement

Unit-based measurement starts by defining the ideal properties of the
measurement, and derives a procedure to aggregate the responses on
items into an overall score that will meet the ideal.

Section \@ref(sec:questions) highlighted questions for individuals,
groups and populations. There are basically three questions:

* What is the difference in development over time for the same child,
group or population?

* What is the difference in development between different children,
groups or populations of the same age?

* How does child development compare to a norm?

In the ideal situation, we would like to have a continuous (latent)
variable $D$ (for development) that measures child development. The
scale should allow us to quantify *ability* of persons, groups or
populations from low to high. The scale should have a *constant unit*,
so that a given difference in ability quantifies means the same across
the entire scale, just like that a difference of 10 cm means the same
irrespective of whether we compare the heights of books or houses.
When are these conditions are met, we say that we measure on an
*interval scale*.

If we succeed in creating an interval scale for child development, an
enormous arsenal of technique developed for quantitative variables
opens up to measure, track and analyze child development. We may then
evaluate the development of a child in terms of $D$ points gained,
create age-dependent charts for development (just like growth charts
for height and weight), devise age-conditional measures for child
development, and intelligent adaptive testing schemes. Promising
studies on Dutch data (@jacobusse2006, @jacobusse2007, @vanbuuren2014)
suggest that such benefits are well within reach.

### Example of unit-based measurement

### Limitations of unit-based measurement

## Why unit-based measuremeny {#sec:whyunit}

This chapter reviewed *age-based*, *score-based* and *unit-based*
approaches to the measurement of child development.

Age-based measurement expresses development in age equivalents, whose
precise definition depends on the reference population.

Score-based measurement quantifies development by summing the number
of passes, but different instruments make different selections of
items, so the measurements are unique to the instruments, and cannot
be compared across instruments.

What we need is the definition of a unit for child development. If we
success in this, we are able to construct different instruments that
yields measurement in a common metric. The next chapter introduces
the basic concepts for the unit-based approach.
