# $D$-score: A unit to quantify general child development {#ch:newmodel}

Chapter \@ref(ch:history) provided historic background on the nature
of child development. Chapter \@ref(ch:threeways) discussed three
general quantification approaches. This chapter explains the concepts
for the unit-based approach, and how to arrive at the $D$-score scale.
The text uses Dutch data to illustration the process.

## The Dutch Development Instrument (DDI) {#sec:ddi}

### Setting

The Dutch Youth Health Care (YHC) routinely monitors development of
almost all children living in The Netherlands. During the first four
years, there are 13 scheduled visits. During these visits, the YHC
professionals evaluate growth and development of the child.

The *Dutch Development Instrument* (DDI; in Dutch: *Van Wiechenschema*) 
is the standard instrument used to measure development
during the ages 0-4 years. The DDI consists of 
[75 milestones](https://www.ncj.nl/van-wiechen/kenmerken/). 
The instrument assesses three developmental domains:

1. Fine motor, adaptation, personality and social behavior;
2. Communication
3. Gross motor

The milestones are organized on two
[forms](https://www.ggdghorkennisnet.nl/?file=656&m=1310474916&action=file.download),
one for children aged 0-15 months, and another for children ages 15-54
months. The YHC professionals administer an age-appropriate subset at
each of the scheduled visits, thus building a 
*longitudinal developmental profile* for each child.

### Description of SMOCC study {#sec:smocc}

The Social Medical Survey of Children Attending Child Health Clinics
(SMOCC) study is a a nationally representative cohort of 2,151
children born in The Netherlands during the years 1988–1989
[@herngreen1994]. The study monitored child development using
observations made on the DDI during nine visits covering the first 24
months of life. The SMOCC study collected information during the first
two years on 57 (out of 75) milestones.

The *standard* set in the DDI consists of relatively easy milestones
that 90 percent of the children can pass at the scheduled age. This
set of designed to have maximal sensitivity for picking up delays in
development. A special feature of the SMOCC study was the inclusion of
additional, and more difficult, milestones beyond the standard set.
The *additional* milestones are taken as the set for the next time
point. The success percentage on additional milestones is about 50
percent.

### Codebook of DDI 0-30 months

```{r smoccitems, echo = FALSE}
library(gseddata)
meta <- dmetric::get_itemtable(get_itemnames(instrument = "ddi")) %>%
  dplyr::select(item, domain, label) %>%
  mutate(domain = plyr::revalue(domain, c("cm" = "Communication", 
                                          "fm" = "Fine motor",
                                          "gm" = "Gross motor")))
data <- get_data(cohort = 53)
df <- data.frame(
  item = get_itemnames(data),
  debut = c(1, 2, 3, 5, 6, 6, 7, 8, 9, 10, 10,
                 4, 7, 8, 9,
                 1, 2, 3, 3, 4, 5, 5, 5, 6, 7, 7, 8, 8, 9, 9, 10, 10, 10, 10, 
                 4, 1, 1, 3, 4, 1, 3, 4, 4, 5, 5, 5, 6, 6, 6, 7, 7, 8, 8, 9, 10, 9, 9), 
  stringsAsFactors = FALSE)
meta <- df %>%  
  left_join(meta, by = "item") %>% 
  mutate(
    itemorder = order_itemnames(item, order = "idnm"), 
    debut = factor(debut, 
                   labels = c("1m", "2m", "3m", "6m", "9m", "12m", "15m", "18m", "24m", "30m"))
    ) %>% 
  arrange(debut, itemorder) %>% 
  dplyr::select(item, debut, domain, label)
  
# calculate visit from data
# meta <- data$itm %>%
#   group_by(item) %>% 
#   summarize(min_age = min(agedays, na.rm = TRUE)) %>% 
#   mutate(visit = cut(min_age, 
#                     breaks = c(18, 43, 76, 149, 216, 320, 399, 481, 633, 800),
#                     labels = c("1m", "2m", "3m", "6m", "9m", "12m", "15m", "18m", "24m"),
#                     right = FALSE),
#          wave = as.integer(visit)) %>% 
#   arrange(wave) %>% 
#   dplyr::select(item, visit) %>% 
#   left_join(meta, by = "item")

options(knitr.table.format = "html") 
kable(meta, 
      caption = "Codebook of DDI as used in the SMOCC study",
      row.names = FALSE, 
      col.names = c("Item", "Debut", "Domain", "Label")) %>%
  kableExtra::kable_styling() %>%
  kableExtra::scroll_box(width = "100%", height = "400px")
```

Table \@ref(tab:smoccitems) shows the 57 milestones from the DDI for
ages 0 - 30 months as they were administered in the SMOCC study. Items
are sorted according to *debut*, the age at which the item appears in
the DDI. The response to each milestones is either a PASS (1) or a
FAIL (0). Children who did not pass a milestone at the debut age were
re-measured on that milestone during the next visit. This process was
continued until a PASS was recorded.

## Probability of passing a milestone given age

```{r smoccfit, echo = FALSE, warning = FALSE}
items <- get_itemnames(data)
varlist <- list(adm = c("cohort", "subjid", "agedays"), 
                items = items,
                cov = NULL)
model <- dmetric::fit_dmodel(varlist, data, name = "57_0")
```

```{r smoccpa, echo = FALSE, fig.cap = '(ref:smoccpa)', fig.height = 4}
# define data for rug plot
data_rug <- data$itm %>%
  mutate(agemos = round(12 * agedays / 365.25), 3) %>%
  dplyr::select(item, value, agemos)

# calculate summary statistics
pass <- data_rug %>%
  mutate(agegp = cut(.data$agemos, breaks = seq(0, 60, 1))) %>%
  group_by(item, agegp) %>%
  summarise(p = round(100 * mean(.data$value, na.rm = TRUE)),
            a = mean(.data$agemos, na.rm = TRUE),
            n = n()) %>%
  ungroup() %>%
  left_join(dmetric::get_itemtable(items = items, package = "gseddata"), by = "item") %>% 
  filter(n >= 5) %>%
  arrange(.data$item, a) %>%
  mutate(domain = plyr::revalue(domain, c("cm" = "Communication", 
                                          "fm" = "Fine motor",
                                          "gm" = "Gross motor")))

plot <- ggplot(pass, aes(a, p, group = item, colour = domain,
                         label = label)) +
  scale_x_continuous("Age (in months)", limits = c(0, 30),
                     breaks = seq(0, 28, 4)) +
  scale_y_continuous("% pass", breaks = seq(0, 100, 20),
                     limits = c(0, 100)) + 
  theme_light() +
  geom_line(size = 0.3) + 
  geom_point(size = 1.0)  +
  theme(legend.position = "none")
py <- plotly::ggplotly(plot, tooltip = c("item", "label"))
py
```

(ref:smoccpa) Empirical percentage of passing each milestone in the
DDI against age (Source: SMOCC data, $n$ = 2151, 9 occasions).

Figure \@ref(fig:smoccpa) summarizes the response obtained on each
milestone as a curve against age. The percentage of pass scores
increases with age for all milestones. Note that curves on the left
have steeper slopes than those on the right, thus indicating that
development is faster for younger children.

The item are colored by domain. In general, domains are well mixed
across age, though around some ages, e.g., at 4 months, multiple
milestones from the same domain appear.

## Probability of passing a milestone given $D$-score

```{r smoccpd, echo = FALSE, fig.cap = '(ref:smoccpd)', fig.height = 4}
modelb <- model$dscore
delta <- dscore::gettau(model$items, model$itembank, lexicon = model$lexicon)
data2 <- data$itm %>%
  filter(item %in% items) %>%
  left_join(data$visit, by = c("subjid", "agedays")) %>%
  left_join(modelb, by = c("subjid", "agedays")) %>%
  dplyr::select(subjid, agedays, item, value, b)

pass <- data2 %>% 
  tidyr::drop_na("value", "b") %>%
  mutate(dgp = cut(.data$b, breaks = seq(0, 70, 1)),
         agemos = round(.data$agedays / 365.25 * 12, 3)) %>%
  group_by(dgp, item) %>%
  summarise(
    p = round(100 * mean(.data$value, na.rm = TRUE)),
    a = mean(.data$agemos, na.rm = TRUE),
    b = mean(.data$b, na.rm = TRUE),
    n = n())
pass <- pass %>%
  group_by() %>%
  left_join(model$itemtable, by = "item") %>%
  arrange(.data$item, dgp) %>%
  dplyr::select(item, dgp, a, b, p, n, domain, label) %>%
  filter(n >= 10 & p > 0 & p < 100) %>%
  mutate(domain = plyr::revalue(domain, c("cm" = "Communication", 
                                          "fm" = "Fine motor",
                                          "gm" = "Gross motor")))

plot <- ggplot(pass, aes(b, p, group = item, colour = domain, 
                         label = label)) +
  scale_x_continuous(paste0("D-score (", model$name,")"),
                     limits = c(0, 70),
                     breaks = seq(0, 70, 10)) +
  scale_y_continuous("% pass", breaks = seq(0, 100, 10),
                     limits = c(0, 100)) +
  theme_light() +
  geom_line(size = 0.3) + 
  geom_point(size = 1.0) +
  theme(legend.position = "none")
py <- plotly::ggplotly(plot, tooltip = c("item", "label"))
py
```

(ref:smoccpd) Empirical percentage of passing each milestone in the
DDI against the $D$-score (Source: SMOCC data, $n$ = 2151, 9 occasions).

Figure \@ref(fig:smoccpd) is similar to Figure \@ref(fig:smoccpa), but
with the horizontal axis replaced by the $D$-score. The $D$-score
summarizes development into one number. See
\@ref(sec:dscoreestimation) for a detailed explanation on how to
calculate the $D$-score. The vertical axis with percent pass is
unchanged.

The percentage of successes increases with $D$-score for all
milestones. In contrast to Figure \@ref(fig:smoccpa) all curves have a
similar slope, a desirable property needed for an interval scale with
a constant unit of measurement (c.f. Section \@ref(sec:unitbased)).

How can the relation between percent pass and age be so different
from the relation between percent pass and the $D$-score? The next
section provides the explanation.

## Relation between age and the $D$-score

```{r smoccda, echo = FALSE, fig.cap = '(ref:smoccda)', fig.height = 6, warning = FALSE}
reference <- dplyr::select(dscore::Dreference, month, SDM2:SDP2) %>%
  filter(month <= 30) %>%
  tidyr::gather(key = centile, value = d,-month)
model$dscore$agemos <- round(model$dscore$agedays / 365.25 * 12, 3)
plot <-
  ggplot(reference, aes(x = month, y = d, group = centile)) +
  theme_light() +
#  coord_fixed(ratio = 60/80 * 688/818) +
  scale_colour_manual(values = dmetric::get_palette("study", package = model$data_package),
                      na.value = "grey") +
  scale_x_continuous("Age (in months)",
                     limits = c(0, 30),
                     breaks = seq(0, 30, 6)) +
  scale_y_continuous(
    paste0("D-score (", model$name, ")"),
    breaks = seq(0, 70, 10),
    limits = c(-5, 75)
  ) +
# geom_line(colour = get_palette("study", package = model$data_package)["GCDG-NLD-SMOCC"]) +
  geom_line(colour = "grey90") +
  geom_point(
    mapping = aes(
      x = agemos,
      y = b,
      group = cohort,
      colour = cohort
    ),
    data = model$dscore,
    size = 0.5,
    shape = 1
  ) +
  facet_wrap( ~ cohort, ncol = 4) +
  theme(legend.position = "none")
plot
```

(ref:smoccda) Relation between child $D$-score and child age in a
cohort of Dutch children (Source: SMOCC data, $n$ = 2151, 9
occasions).

The relation between $D$-score and age is nonlinear. Development in
the first year is more rapid than in the second year. During the first
year infants gain about 40$D$, whereas in the second year they gain
about 20$D$. A similar change in growth rate occurs in length (first
year: 23 cm, second year: 12 cm, for Dutch children).

```{r smoccdap, echo = FALSE, fig.cap = '(ref:smoccdap)', warning = FALSE}
# model ICC curves D-score
grid <- crossing(
  item = items,
  b = seq(0, 70, 0.2))
ib <- model$itembank %>%
  rename(item = lex_gsed) %>%
  dplyr::select(item, domain, label, tau)
tab <- grid %>%
  left_join(ib, by = "item") %>%
  mutate(p = 100 * plogis(b, location = tau, scale = model$transform[2])) %>%
  filter(p > 1 & p < 99) %>%
  mutate(a = approx(x = dscore::Dreference$mu, 
                    y = dscore::Dreference$month, xout = b)$y) %>%
  tidyr::drop_na("a") %>%
  group_by(item)

py <- plot_ly(tab, x = ~a, y = ~b, z = ~p, color = ~domain) %>%
  add_lines() %>%
  layout(scene = list(xaxis = list(title = 'Age'),
                      yaxis = list(title = 'D-score'),
                      zaxis = list(title = 'Probability'),
                      aspectratio = list(x = 1, y = 1, z = 0.5),
                      camera = list(
                        up = list(
                          x = 1,
                          y = 0,
                          z = 0
                        ),
                        eye = list(
                          x = 0,
                          y = 0,
                          z = 2
                        )
                      )
  ))
py
```

(ref:smoccdap) 3D-line graph illustrating how the patterns in Figures \@ref(fig:smoccpa) and \@ref(fig:smoccpd) induce the curvature in the relation between $D$-score and age.

Figure \@ref(fig:smoccdap) shows the mutual relations between age,
milestone success and the $D$-score.

* Rotate the graph to the age - probability plane, and observe the
commonality with the pattern in Figure \@ref(fig:smoccpa) with unequal
slopes; 

* Rotate the graph to the $D$-score - probability plane, and observe
the commonality with the pattern in Figure \@ref(fig:smoccpd) with
equal slopes;

* Rotate the graph to the $D$-score - age plan, and observe that all
patterns can co-exist because of the curvature in the relation between
$D$-score and age.

## Measurement model for the $D$-score {#sec:measurementmodel}

### What are measurement models? 

From section \@ref(sec:whatismeasurement) we quote:

> The measurement model specifies the relations between the data 
> and the latent variable.

The scientific theory of measurement models is known under the name of
*Item Response Theory* (IRT). Good introductory works include
@wright1982, @embretsen2000 and @engelhard2013.

IRT models enables quantification of the locations of both *items*
(milestones) and *persons* on the latent variable. We reserve the term
*item* for generic properties, and *milestone* for child development.
In general, items are part of the measurement instrument, persons are
the objects to be measured.

An IRT model has three major structural components: 
 
* Specification of the underlying *latent variable(s)*. In this
work, we restrict ourselves to models with just one latent variable.
Multi-dimensional IRT models do have their uses, but they are
complicated to fit and not widely used;

* For a given item, a specification of the *probability of success*
given a value on the latent variables. This specification can take
many forms. Section \@ref(sec:itemresponsefunctions) focuses on this
in more detail;

* Specification how probability models for the different items should
be combined. In this work, we will restrict to models that assume
*local independence* of the probabilities. In that case, the
probability that two items are passed can be found by multiplying each
item's success probability.

### Adapt the model? Or adapt the data? {#sec:adaptmodel}

The measurement model induces a predictable pattern in the observed
items, and this pattern can be tested against the observed data. When
there is misfit between the expected and observed data, we can follow
two strategies:

* Make the measurement model more general.

* Discard items (and sometimes persons) to make the model fit.

These are very different strategies that have led to heated debates
among psychometricians. See @engelhard2013 for an overview.

In this work, we opt for the - very strict - Rasch model (@rasch1960),
and will adapt the data to reduce discrepancies between model and
data. Arguments for this choice are given later, in Section
\@ref(sec:whyrasch).

## Item response functions {#sec:itemresponsefunctions}

Most measurement models describe the probability of passing an item as
a function of the *difference* between the person's ability and the
item's difficulty. A person with low ability will almost surely fail a
difficult item, whereas a highly able person will almost surely pass
an easy item.

Let us now introduce a few symbols. We adopt the notation used in
@wright1982. See Chapter \@ref(ch:notation) for a complete list.

Symbol        | Term          | Description
------------- | ------------- | -----------
$\beta_n$     | Ability       | True (but unknown) developmental score of child $n$
$\delta_i$    | Difficulty    | True (but unknown) difficulty of an item $i$
$\pi_{ni}$    | Probability   | Probability that child $n$ passes item $i$

Thus, Figure \@ref(fig:lineplot) shows three $\beta_n$'s and four
$\delta_i$'s.

The difference between ability of child $n$ and difficulty of item $i$ is

$$\beta_n - \delta_i$$

In the special case that $\beta_n = \delta_i$, the person will have a
probability of 0.5 of passing the item. 

### Logistic model

A widely used method to map differences on the latent scale are
*logistic units* (or  *logits*) [@berkson1944]. In `R` we can
calculate the probability by the inverse-logit (or *logistic*
function) for a difference of 0 and 1 as

```{r logit0}
plogis(c(0, 1))
```

```{r logisticplot, echo = FALSE, warning = FALSE, fig.asp = 0.5, fig.cap = '(ref:logisticplot)', width = "80%"}
pkg <- c("ggplot2", "RColorBrewer")
loaded <- sapply(pkg, require, character.only = TRUE, 
                 warn.conflicts = FALSE, quietly = TRUE)
options(knitr.kable.NA = "?")

x <- seq(-5, 5, 0.25)
items <- data.frame(
  Items = c(rep(c("C", "B", "A", "D", "E"), times = c(4, 2, length(x), length(x), length(x)))),
  beta = c(-5, 1, 1, +5,  -5,  +5, x, x, x),
  pass = c( 0, 0, 1,  1, 0.3, 0.3,
           plogis(x),
           plogis(0.6 * (x + 1)),
           smooth.spline(x = x, y = plogis(c(x[1:15], rep(x[16], 10), x[26:41])), df = 8)$y))
ggplot(subset(items, Items == "A"), aes(x = beta, y = 100 * pass, group = Items, colour = Items)) +
  geom_line(lwd = 0.55) + 
  scale_x_continuous("logits", breaks = -5:5) + 
  scale_y_continuous("% pass", breaks = seq(0, 100, 20), limits = c(0, 100)) +
  scale_colour_brewer(palette = "Set1", labels = c("A: Logistic", "B: Constant", "C: Step", "D: 2PL", "E: Monotone")) + 
  theme_light() + 
  theme(legend.position = c(0.05, 0.95), legend.justification = c(0, 1))
```

(ref:logisticplot) Standard logistic curve. Percentage of children
passing an item for a given ability-difficulty gap $\beta_n -
\delta_i$.

Figure \@ref(fig:logisticplot) shows how the percentage that pass the
item varies in terms of the ability-difficulty gap $\beta_n -
\delta_i$. The gap  can vary either by $\beta_n$ or $\delta_i$, so we
may use the graph in two ways:

* To find the probability of passing various items for a child with
ability $\beta_n$. If $\delta_i = \beta_n$ then $\pi_{ni} = 0.5$. If
$\delta_i < \beta_n$ then $\pi_{ni} < 0.5$, and if $\delta_i >
\beta_n$ then $\pi_{ni} > 0.5$.

* To find the probability of passing a given item $\delta_i$ for
children that vary in ability. If $\beta_n = \delta_i$ then $\pi_{ni} =
0.5$. If $\beta_n < \delta_i$ then $\pi_{ni} < 0.5$, and if $\beta_n >
\delta_i$ then $\pi_{ni} > 0.5$.

### Types of item response functions

The standard logistic function is by no means the only option to map
the relation between the latent variable and the probability of
passing an item. The logistic function is the dominant choice in IRT,
but it is instructive to study some other mappings. The *item response
function* maps success probability against ability.

```{r irfplot, echo = FALSE, warning = FALSE, fig.asp = 0.5, fig.cap = '(ref:irfplot)'}
ggplot(items, aes(x = beta, y = 100 * pass, group = Items, colour = Items)) +
  geom_line(lwd = 0.75) + 
  scale_x_continuous("Child ability (logits)", breaks = -5:5) + 
  scale_y_continuous("% pass", breaks = seq(0, 100, 20), limits = c(0, 100)) +
  scale_colour_brewer(palette = "Set1", labels = c("A: Logistic", "B: Constant", "C: Step", "D: 2PL", "E: Monotone")) + 
  theme_light() + 
  theme(legend.position = c(0.05, 0.95), legend.justification = c(0, 1))
```

(ref:irfplot) Item response functions for five hypothetical items,
each demonstrating a positive relation between ability and probability
to pass.

Figure \@ref(fig:irfplot) illustrates several other possibilities. Let
us consider five hypothetical items, A-E. Note that the horizontal
axis now refers to ability, instead of ability-item gap in
\@ref(fig:logisticplot).

- A: Item A is the logistic function discussed in Section
\@ref(sec:itemresponsefunctions).

- B: For item B, the probability of passing is constant at 30 percent.
This 30 percent is not related to ability. This item does not measure
ability, and only adds to the noise. It is a bad item for measurement.

- C: Item C is step function centered at an ability level of 1, so *all*
children with an ability below 1 logit fail, and *all* children with
ability above 1 pass. This is the ideal item for discriminating
children with an abilities above and below 1. The item is not
sensitive to differences at other ability levels, and often not so
realistic in practice.

- D: Like A, item D is a smoothly increasing logistic function, but it
has an extra parameter that allows it to vary its slope (or
discrimination). The extra parameter can make the curve steeper (more
discriminatory) than the red curve, in the limit approaching a step
curve. It can also become shallower (less discriminatory) than the red
curve (as plotted here), in the limit approaching a constant curve
(item B). Thus, item D generalizes items A, B or C.

- E: Item E is even more general in the sense that it need not be
logistic, but a general monotonically increasing function. As plotted,
the item is insensitive to abilities between -1 and 0 logits, and more
sensitive to abilities between 0 to 2 logits.

These are just some examples of how the relation between the child's
ability and passing probability could look. In practice, the curves
need not start at 0 percent or end at 100 percent. They could also be
U-shaped, of have other non-monotonic forms. See @coombs1964 for a
thorough overview of such models. In practice, most models are
restricted to shapes A-D.

### Person response functions

The roles of persons and items can be switched. The *person response
function* tells us how likely it is that a single person can pass an
item, or more commonly, a set of items.

Let us continue with items A, C and D from Figure \@ref(fig:irfplot), and 
calculate the response function for three children, respectively with 
abilities $\beta_1 = -2$, $\beta_2 = 0$ and $\beta_3 = 2$.

```{r prfplot, echo = FALSE, warning = FALSE, fig.asp = 0.5, fig.cap = '(ref:prfplot)'}
x <- c(seq(-5.00, -2.25, 0.25), -2.001, -2, -1.999, 
       seq(-1.75, -0.25, 0.25), -0.001,  0,  0.001,
       seq( 0.25,  1.75, 0.25),  1.999,  2,  2.001,
       seq( 2.25,  5.00, 0.25))
A1 <- c(rep(1, 23), 0.5, rep(0, 23))
C1 <- 1 - plogis(x)
D1 <- 1 - plogis(0.6 * (x + 1))
ACD1 <- (A1 + C1 + D1) / 3

A2 <- c(rep(1, 33), 0.5, rep(0, 13))
C2 <- 1 - plogis(x - 2)
D2 <- 1 - plogis(0.6 * (x - 1))
ACD2 <- (A2 + C2 + D2) / 3

A3 <- c(rep(1, 13), 0.5, rep(0, 33))
C3 <- 1 - plogis(x + 3)
D3 <- 1 - plogis(0.6 * (x + 4))
ACD3 <- (A3 + C3 + D3) / 3

data <- data.frame(delta = x,
                   person = rep(c("-2", "0", "+2"), each = 47),
                   pass = c(ACD3, ACD1, ACD2))
ggplot(data, aes(x = delta, y = 100 * pass, group = person, colour = person)) +
  geom_line(lwd = 0.75) + 
  scale_x_continuous("Item difficulty (logits)", breaks = -5:5) + 
  scale_y_continuous("% pass", breaks = seq(0, 100, 20), limits = c(0, 100)) +
  scale_colour_brewer(name = "Person ability", palette = "Set1", limits = c("-2", "0", "+2")) + 
  theme_light() + 
  theme(legend.position = c(0.8, 0.6), legend.justification = c(0, 0)) +
  annotate("text", x = -4.5, y = 15, label = "Easy items", fontface = "bold") + 
  annotate("text", x =  4.5, y = 15, label = "Hard items", fontface = "bold")
```

(ref:prfplot) Person response functions for three children with abilities 
-2, 0 and +2, using a small test of items A, C and D.

Figure \@ref(fig:prfplot) presents the person response functions from
three persons with abilities of -2, 0 and +2 logits. The functions are
calculated as the average of response probabilities on items A, C and
D. Thus, on average, we expect that child 1 will pass an easy item of
difficulty -3 in about 60 percent of the time, whereas for an
intermediate item of difficulty of -1 the passing probability would be
10 percent. For child 3, with a higher ability, these probabilities
are quite different: 97% and 90%. The substantial drop in the middle
of the curve is caused by the step function of item A. 

## Engelhard criteria for invariant measurement {#sec:engelhard}

In this work, we strive to achieve *invariant measurement*, a very
strict form of measurements that is subject to the following
requirements (@engelhard2013, p. 14):

1. *Item-invariant measurement of persons*: The measurement of persons
must be independent of the particular items that happen to be used for
the measuring.

2. *Non-crossing persons response functions*: A more able person must
always have a better chance of success on an item that a less able
person.

3. *Person-invariant calibration of test items*: The calibration of
the items must be independent of the particular persons used for
calibration.

4. *Non-crossing item response functions*: Any person must have a
better chance of success on an easy item than on a more difficult
item.

5. *Unidimensionality*: Items and persons must be simultaneously
located on a single underlying latent variable.

Three families of IRT models actually support invariant measurement:

1. Scalogram model (@guttman1950)

2. Rasch model (@rasch1960, @andrich1978, @wright1982)

3. Mokken scaling model (@mokken1971, @molenaar1997)

The Guttman and Mokken model yield an ordinal latent scale, while the 
Rasch model yields an interval scale (with a constant unit).

## Why take the Rasch model? {#sec:whyrasch}

* *Invariant measurement*: The Rasch model meets the five Engelhard
criteria (c.f. Section \@ref(sec:engelhard)).

* *Interval scale*: When it fits, the Rasch model provides an interval
scale, the de-facto requirement for any numerical comparisons (c.f.
Section \@ref(sec:motivationunit)).

* *Parsimonious*: The Rasch model has one parameter for each item and
one parameter for each person. It's one of the most parsimonious IRT
models, and can easily be applied to thousands of items and millions
of persons.

* *Specific objectivity*: Person and item parameters are
mathematically separate entities in the Rasch model. In practice, this
means that the estimated difference in ability between two persons
does not depend on the difficulty of the test. Also, the estimated
differences in difficulties between two items does not depend on the
abilities in the calibration sample.

* *Unified model*: The Rasch model unifies distinct traditions in
measurement theory. The Rasch model can be derived from
    + [Thorndike's 1904 criteria](https://www.rasch.org/rmt/rmt143g.htm) 
    + [Guttman scalogram model](https://www.rasch.org/rmt/rmt63e.htm)
    + [Ratio-scale counts](https://www.rasch.org/rmt/rmt62c.htm)
    + [Raw scores as sufficient statistics](https://www.rasch.org/rmt/rmt32e.htm)
    + [Thurstone's scaling requirements](https://www.rasch.org/rmt/rmt21a.htm)
    + [Campbell concatenation](https://www.rasch.org/rmt/rmt21b.htm)
    + [Rasch's specific objectivity](https://www.rasch.org/rmt/rmt11a.htm)


* *Fits child development data*: Last but not least, as we will see
in Section \@ref(sec:fitddi), the Rasch model provides excellent fit to
child development milestones.
