# Computation {#ch:computation}

This chapter explains the basic computations needed for fitting and
evaluation the Rasch model. We distinguish the following steps:

* Data preparation (\@ref(sec:datapreparation))
* Estimation of item parameters (\@ref(sec:itemestimation))
* Anchoring (\@ref(sec:anchoring))
* Estimation of the $D$-score (\@ref(sec:dscoreestimation))

Reader not interested in these details may continue to model
evaluation in Chapter \@ref(ch:evaluation).

## Data preparation {#sec:datapreparation}

Data SMOCC data of Secion \@ref(sec:ddi)

## Item parameter estimation {#sec:itemestimation}

There are many computer routines for this task, but only a few produce
the desired result when applied to developmental data.

```{r smoccfit2, echo = FALSE, warning = FALSE}
library(gseddata)
data <- get_data(cohorts = 53)
items <- get_itemnames(data)
varlist <- list(adm = c("cohort", "subjid", "agedays"), 
                items = items,
                cov = NULL)
model <- fit_dmodel(varlist, data, name = "57_0")
```


## Anchoring {#sec:anchoring}



## Estimation of the $D$-score {#sec:dscoreestimation}

The $D$-score quantifies development of a child. In order to calculate
the $D$-score we need the following ingredients:

* Child's PASS/FAIL scores on the milestones administered;

* The difficulty estimates of each milestone administered;

* A prior distribution, an estimate of the $D$-score distribution
before seeing any PASS/FAIL score.

Using these inputs, we may use Bayes theorem to calculate the position
of the person on the latent variable. The method, first described by
@bock1982, an is known as the Expected a posteriori (EAP) estimation.

### EAP algorithm

Calculation of the $D$-score proceeds item by item. Suppose we have
some vague and preliminary idea about the distribution of $D$. The
procedure uses Bayes rule to update this prior knowledge with data
from the first item (fail/pass score, item difficulty) to calculate
the posterior. In the next step, this posterior is used as the prior
for the next item, and so on. The procedure stops when all items have
been used. The order in which items are entered does not matter for
the end result. The $D$-score is equal to the mean of the last
posterior.

### Age-dependent starting prior

In the first step, we need a starting distribution as prior. The
starting prior should define the range of reasonable values for $D$.
For example, an infant aged 2 months the $D$-score will have lower
$D$-score than a baby aged 12 months. The prior should be a bit
informative but not too much, so it was decided to choose a broad
distribution which was centred quite well. The prior was chosen as a
normal distribution with a mean equal to the $D$-score expected at the
child's age, and with a standard deviation of 5, which about twice the
normal variation in the $D$-score.

```{r figpriors, echo = FALSE,  fig.cap="\\label{fig:figpriors} Age-dependent starting priors for the D-score at the ages of 1, 15 and 24 months", fig.height = 4}
library(dscore)
library(ggplot2)
#example from the adp function in ggplot:
qp <- -10:80

p1m <- data.frame(qp = qp, month = rep(1, length(qp)), p = adp(1/12, qp))
p15m <- data.frame(qp = qp, month = rep(15, length(qp)), p = adp(15/12, qp))
p24m <- data.frame(qp = qp, month = rep(24, length(qp)), p = adp(24/12, qp))

dataprior <- rbind(p1m, p15m, p24m) %>%
  mutate(month = as.factor(month)) %>%
  dplyr::filter(p > 0.00001)

p <- ggplot(dataprior, aes(qp, p, group = month)) +
  theme_light() + 
  theme(legend.position = c(0.05, 0.95), legend.justification = c(0, 1)) +
  geom_path(aes(col = month)) +
  coord_cartesian(xlim = c(-10, 80)) +
  scale_y_continuous(name = "Density", limits = c(0, 0.3)) +
  scale_x_continuous(name = "D-score") +
  labs(fill = "Age (in months)")
p
```

Figure \@ref(fig:figpriors) illustrates starting distributions
(priors) for the ages of 1, 15 and 24 months. As expected, the assumed
ability of an infant aged of 1 month is lower than that of a child
aged 15 months, which in turn is lower than the ability of a toddler
aged 24 months.

### Numerical example

Suppose we measure two boys aged 15 months, David and Rob, by the DDI.
David passes the first four milestones, but does not complete the test.
Rob completes the test, but fails on two out of five milestones. How 
do we calculate the $D$-score for David and Rob?

```{r datadavidandrob, echo = FALSE}
items <- c("ddifmd011", "ddifmm012", "ddicmm037", "ddigmm066", "ddigmm067")
ib <- model$itembank %>%
  rename(item  = lex_gsed, delta = tau) %>%
  filter(item %in% items) %>%
  dplyr::select(item, label, delta)
tab <- data.frame(ib, 
                  David = c(1, 1, 1, 1, NA),
                  Rob = c(1, 0, 1, 1, 0))
kable(tab, caption = "Scores of David and Rob on five milestones from the DDI", 
      row.names = FALSE, col.names = c("Item", "Label", "Delta", "David", "Rob"))
```


First, it is important that the names of the items used are similar to
the names of the items in the built-in item bank. In this example the
lexicon *gcdg* is used. Moreover, the data should be in long format
with the following columnnames: *items*, *scores* and *ages*. This
leads to table \@ref(tab:tablenewWS1).

```{r, tablenewWS1, echo=FALSE}
WS1 <- data.frame(child = c("child 1", "child 1", "child 1", "child 1", "child 1", "child 1", "child 2", "child 2", "child 2", "child 2", "child 2", "child 2"),items = c("n32", "n33", "v38", "n37", "n34", "n35", "n32", "n33", "v38", "n37", "n34", "n35"), scores = c(1, 1, 1, 1, 1, 1, 1, 0, 1, 1, 1, 0), ages = c(15/12, 15/12, 15/12, 15/12, 15/12, 15/12, 15/12, 15/12, 15/12, 15/12, 15/12, 15/12))

kable(WS1, caption = "Long format of the example of scores on the items of the DDI at the age of 15 months", row.names = NA)
```

Now that the data is in long format and the item names are changed to
the lexicon *gcdg*, we can calculate the D-score for both children. In
the first step of the iterative procedure, we combine the
age-dependent prior of 15 months with the score of the first item
*n34* to define the posterior. Item *n34* is a gross motor item:
'crawls, abdomen off the floor'. In table \@ref(tab:tablenewWS1), we
can see that both example children were able to crawl with their
abdomen off the floor. Note, that we could have chosen any of the
items as the first item, since the D-score is independent of the order
of items in the calculation.

The age dependent prior of 15 months and the posterior distribution
after adding information from item *n34* are given in figure
\@ref(fig:fig1item). In this figure, we can see that the posterior
distribution doesnot differ that much from the prior distribution.
Both distributions are centred at the same point, while the posterior
distribution is smaller than the prior distribution indicating more
precision. Since both example children were able to crawl with their
abdomen off the floor, the posterior distribution after the first step
of the procedure is equal.

```{r, fig1item, echo=FALSE, fig.cap="\\label{fig:fig1item} Age-dependent priors of 15 months and posterior after 1st item n34"}
items <- c("n34")
age <- 1.25
  
#full posterior 1e item
qp <- -10:80
fp <- dscore(1, items, age, full = TRUE, lexicon = "gcdg", qp = qp)
fp1 <- data.frame(qp = qp, fp = fp[[1]]$posterior, distribution = rep("posterior", length(qp)))
prior <- data.frame(qp = qp, fp = fp[[1]]$start, distribution = rep("prior", length(qp)))

dataprior <- rbind(fp1, prior)
dataprior$distribution <- as.factor(dataprior$distribution)

p1 <- ggplot(dataprior, aes(qp, fp, group = distribution)) +
      geom_path(aes(col = distribution)) +
      coord_cartesian(xlim=c(-10, 80)) +
      scale_y_continuous(name = "Density") +
      scale_x_discrete(name = "D-score")
p1
```

Now, let's add an item on which the scores for the two example
children differ, e.g. item *n35*. Item *n35* is also a gross motor
item, namely 'walks along'. So both children are able to crawl (with
their abdomen of the floor), but only *child 1* is able to walk along.
We can use the previously calculated posterior distribution as prior
when we add item *n35*.

Let's start with *child 1*, who is able to walk along. Using Bayes
rule, the new posterior distribution can be calculated. In figure
\@ref(fig:fig2itemC1) the prior, in this case the posterior after
adding item *n34* to the age-dependent prior, and the newly calculated
posterior distribution are visualized. Again, the posterior and prior
are centred around the same point, while the posterior has a bit more
precision.

```{r, fig2itemC1, echo=FALSE, fig.cap="\\label{fig:fig2itemC1} Prior and posterior in the 2nd step of the iterative process for child 1"}
items <- c("n34", "n35")
age <- rep(1.25, length(items))
  
#full posterior 2e item
qp <- -10:80
fp2 <- dscore(c(1,1), items, age, full = TRUE, lexicon = "gcdg", qp = qp)
fp1 <- data.frame(qp = qp, fp = fp[[1]]$posterior, distribution = rep("1st posterior", length(qp)))
fp2 <- data.frame(qp = qp, fp = fp2[[1]]$posterior, distribution = rep("2nd posterior", length(qp)))

datafp <- rbind(fp1, fp2)
datafp$distribution <- as.factor(datafp$distribution)

p2 <- ggplot(datafp, aes(qp, fp, group = distribution)) +
      geom_path(aes(col = distribution)) +
      coord_cartesian(xlim=c(-10, 80)) +
      scale_y_continuous(name = "Density") +
      scale_x_discrete(name = "D-score")
p2


```

Now we can do the same thing for *child 2*, who is able to crawl with
his/her abdomen off the floor (*n34*) but not to walk along (*n35*).
Again, we can use the posterior after adding information from *n34* as
the prior and use the information from *n35* to calculate the
posterior. In figure \@ref(fig:fig2itemC2) the prior and posterior are
given. Now the prior and posterior are different, the posterior is on
the left of the prior and has an high precision in comparison to the
prior. Hence, being unable to walk along at the age of 15 months, like
*child 2*, is more indicative for the development of a child of 15
months than being able to walk along like *child 1*.

```{r, fig2itemC2, echo=FALSE, fig.cap="\\label{fig:fig2itemC2} Prior and posterior in the 2nd step of the iterative process for child 2"}
items <- c("n34", "n35")
age <- rep(1.25, length(items))
  
#full posterior 2e item FAIL
qp <- -10:80
fp2 <- dscore(c(1,0), items, age, full = TRUE, lexicon = "gcdg", qp = qp)
fp1 <- data.frame(qp = qp, fp = fp[[1]]$posterior, distribution = rep("1st posterior", length(qp)))
fp2 <- data.frame(qp = qp, fp = fp2[[1]]$posterior, distribution = rep("2nd posterior", length(qp)))

datafp2 <- rbind(fp1, fp2)
datafp2$distribution <- as.factor(datafp2$distribution)

p3 <- ggplot(datafp2, aes(qp, fp, group = distribution)) +
      geom_path(aes(col = distribution)) +
      coord_cartesian(xlim=c(-10, 80)) +
      scale_y_continuous(name = "Density") +
      scale_x_discrete(name = "D-score")
p3

```


The next four steps in the iterative procedure, information from the
other four items (*n32*, *n33*, *v38* and *n37*) can be added in a
similar fashion. Figure \@ref(fig:fig6itemC1C2) shows the final
posterior distributions for *child 1* and *child 2* and the prior
distribution for the age of 15 months. The D-score can now be
determined by looking at the mean of the posterior distributions.
Hence, the D-score for *child 1* and *child 2* are equal to 55.75 and
47.76, respectively.

```{r, fig6itemC1C2, echo=FALSE, fig.cap="\\label{fig:fig6itemC1C2} Prior and final posterior for child 1 and 2"}
items <- c("n32", "n33", "v38", "n37","n34", "n35")
age <- rep(1.25, length(items))
  
#full posterior for both children and prior
qp <- -10:80
fp6_1 <- dscore(c(1,1,1,1,1,1), items, age, full = TRUE, lexicon = "gcdg", qp = qp)
fp6_2 <- dscore(c(1,0,1,1,1,0), items, age, full = TRUE, lexicon = "gcdg", qp = qp)
prior <- data.frame(qp = qp, fp = fp[[1]]$start, distribution = rep("prior", length(qp)))
fp6_1 <- data.frame(qp = qp, fp = fp6_1[[1]]$posterior, distribution = rep("final posterior child 1", length(qp)))
fp6_2 <- data.frame(qp = qp, fp = fp6_2[[1]]$posterior, distribution = rep("final posterior child 2", length(qp)))

datafp6 <- rbind(prior, fp6_1, fp6_2)
datafp6$distribution <- as.factor(datafp6$distribution)

p6 <- ggplot(datafp6, aes(qp, fp, group = distribution)) +
      geom_path(aes(col = distribution)) +
      coord_cartesian(xlim=c(-10, 80)) +
      scale_y_continuous(name = "Density") +
      scale_x_discrete(name = "D-score")
p6

```

```{r, echo=FALSE, message=FALSE}
#Dscore child1
dscore(c(1,1,1,1,1,1), items, age, full = FALSE, lexicon = "gcdg") #55.75
#Dscore child2
dscore(c(1,0,1,1,1,0), items, age, full = FALSE, lexicon = "gcdg") #47.76
```

*nog toe te voegen in de tekst*

*quadratic points*

*Note: met deze methode is het mogelijk om scores voor extremen te berekenen, andere methodes (zoals de worms estimator) kunnen dit niet altijd.*

*Prior die gekozen was: plafond prior (wanneer je alle items kan)*


We already explained how the D-score suffers from a ceiling and bottom
at the start and end of the scale. However, this can also happen
within the D-score scale: some values of the D-score will never occur
no matter the combination of scores on the milestones. This will
mainly occur when calibrated milestones are used to calculate the
D-score. For example, the Dutch Developmental Instrument (DDI), which
is used as a screening instrument, is calibrated: at each wave 90\% of
the children should be able to get a positive score on the milestones.
Which allows for detection of children with a developmental delay.
