# Computation {#ch:computation}

This chapter explains the basic computations needed for fitting and
evaluation the Rasch model. We distinguish the following steps:

* Data preparation (\@ref(sec:datapreparation))
* Estimation of item parameters (\@ref(sec:itemestimation))
* Anchoring (\@ref(sec:anchoring))
* Estimation of the $D$-score (\@ref(sec:dscoreestimation))

Reader not interested in these details may continue to model
evaluation in Chapter \@ref(ch:evaluation).

## Data preparation {#sec:datapreparation}

Data SMOCC data of Secion \@ref(sec:ddi)

## Item parameter estimation {#sec:itemestimation}

There are many computer routines for this task, but only a few produce
the desired result when applied to developmental data.

## Anchoring {#sec:anchoring}



## Estimation of the $D$-score {#sec:dscoreestimation}

Once we know the score on one or more items, and the difficulties of these
items, we may calculate the position of the person on the latent
variable.


With the SMOCC dataset we can compare the D-scores calculated by
making use of calibrated milestones (the current wave milestones) with
the D-scores calculated by making use of non-calibrated milestones
(the next wave milestones). These D-scores for children of the age of
2 and 3 months (wave 2 and 3) are plotted against IQ at 5 years in
figure \@ref(fig:dscoresnw). The continuous variable IQ at 5 years
allows for nice visualization of the variation in D-scores. Moreover,
the densities of the D-score are given at the bottom of the plots
showing the distribution of the values of the D-score.

### Numerical example

To explain how the D-score is calculated and demonstrate the use of
the D-score package two example children, both of the age of 15
months, are used. One of the example children is able to do all items
of his/her age, and the other example child is not able to do all
items of his/her age. In this way, we can show how different scores on
a item affect the D-score.

So for example, for a child of the age of 15 months the Dutch
Developmental Instrument consists of the following six items: puts
cube in and out of a box, plays 'give and take', crawls (abdomen off
the floor), walks along, understands a few words, and says 2
'sound-words' with comprehension. For each of these six items it is
noted whether the child was able to perform this item. An example of
the results can be found in table \@ref(tab:tableWS), in which *child
1* was able to perform all items and *child 2* was able to perform
four of the six items. Based on these scores we can calculate the
dscore for both these children.


So for example, for a child of the age of 15 months the Dutch
Developmental Instrument consists of the following six items: puts
cube in and out of a box, plays 'give and take', crawls (abdomen off
the floor), walks along, understands a few words, and says 2
'sound-words' with comprehension. For each of these six items it is
noted whether the child was able to perform this item. An example of
the results can be found in table \@ref(tab:tableWS), in which *child
1* was able to perform all items and *child 2* was able to perform
four of the six items. Based on these scores we can calculate the
dscore for both these children.

```{r, echo=FALSE}
child1 <- c(15/12, 1, 1, 1, 1, 1, 1)
child2 <- c(15/12, 1, 0, 1, 1, 1, 0)
tableWS <- rbind(child1, child2)
colnames(tableWS) <- c("age in years", "item 1", "item 2", "item 3", "item 4", "item 5", "item 6")
rownames(tableWS) <- c("child 1", "child 2" )
#item2 : speelt geven/nemen
#item6 : loopt langs
```

```{r, tableWS, echo=FALSE}
library(knitr)
kable(tableWS, caption = "Example of scores on the items of the DDI at the age of 15 months", booktabs = TRUE)
```

Calculation of the D-score is an iterative procedure, in which in each
step information of one item is added. The iterative procedure uses
Bayes rule to update the prior (knowledge) with data to calculate a
posterior. In the next step, this posterior is used as the prior and
information, the score, from a new item is added. This results in a
new posterior distribution. When the information of all items have
been added to the procedure, the D-score is equal to the mean of the
final posterior distribution.

*!!plafond prior nog uitleggen!!*

In the first step, the score of the first item is combined with the
prior. However, at the first step we cannot use the previous posterior
as our prior. Hence, we need a starting distribution to use as the
prior in the first step. It is important that this starting
distribution is a bit informative but not too much, so it was decided
to choose a quite broad distribution which was centred quite well. The
prior was chosen in such a way that the mean value of the prior
distribution was equal to the median D-score at that age. The standard
deviation is equal to 5, which about twice the normal variation in the
D-score.

The starting distributions (priors) for the ages of 1, 15 and 24
months are given in figure \@ref(fig:figpriors). This figure shows
that the priors assume that ability of a child of the age of 1 month
is lower than the ability of a child of the age of 15 months, which in
turn is lower than the ability of a child of the age of 24 months.

```{r, figpriors, echo = FALSE,  fig.cap="\\label{fig:figpriors} Priors at ages of 1, 15 and 24 months"}
library(dscore)
library(ggplot2)
#example from the adp function in ggplot:
qp <- -10:80

p1m <- data.frame(qp = qp, month = rep(1, length(qp)), p = adp(1/12, qp))
p15m <- data.frame(qp = qp, month = rep(15, length(qp)), p = adp(15/12, qp))
p24m <- data.frame(qp = qp, month = rep(24, length(qp)), p = adp(24/12, qp))

dataprior <- rbind(p1m, p15m, p24m)
dataprior$month <- as.factor(dataprior$month)

p <- ggplot(dataprior, aes(qp, p, group = month)) +
      geom_path(aes(col = month)) +
      coord_cartesian(xlim=c(-10, 80)) +
      scale_y_continuous(name = "Density") +
      scale_x_discrete(name = "D-score")
p
```

First, it is important that the names of the items used are similar to
the names of the items in the built-in item bank. In this example the
lexicon *gcdg* is used. Moreover, the data should be in long format
with the following columnnames: *items*, *scores* and *ages*. This
leads to table \@ref(tab:tablenewWS1).

```{r, tablenewWS1, echo=FALSE}
WS1 <- data.frame(child = c("child 1", "child 1", "child 1", "child 1", "child 1", "child 1", "child 2", "child 2", "child 2", "child 2", "child 2", "child 2"),items = c("n32", "n33", "v38", "n37", "n34", "n35", "n32", "n33", "v38", "n37", "n34", "n35"), scores = c(1, 1, 1, 1, 1, 1, 1, 0, 1, 1, 1, 0), ages = c(15/12, 15/12, 15/12, 15/12, 15/12, 15/12, 15/12, 15/12, 15/12, 15/12, 15/12, 15/12))

kable(WS1, caption = "Long format of the example of scores on the items of the DDI at the age of 15 months", row.names = NA)
```

Now that the data is in long format and the item names are changed to
the lexicon *gcdg*, we can calculate the D-score for both children. In
the first step of the iterative procedure, we combine the
age-dependent prior of 15 months with the score of the first item
*n34* to define the posterior. Item *n34* is a gross motor item:
'crawls, abdomen off the floor'. In table \@ref(tab:tablenewWS1), we
can see that both example children were able to crawl with their
abdomen off the floor. Note, that we could have chosen any of the
items as the first item, since the D-score is independent of the order
of items in the calculation.

The age dependent prior of 15 months and the posterior distribution
after adding information from item *n34* are given in figure
\@ref(fig:fig1item). In this figure, we can see that the posterior
distribution doesnot differ that much from the prior distribution.
Both distributions are centred at the same point, while the posterior
distribution is smaller than the prior distribution indicating more
precision. Since both example children were able to crawl with their
abdomen off the floor, the posterior distribution after the first step
of the procedure is equal.

```{r, fig1item, echo=FALSE, fig.cap="\\label{fig:fig1item} Age-dependent priors of 15 months and posterior after 1st item n34"}
items <- c("n34")
age <- 1.25
  
#full posterior 1e item
qp <- -10:80
fp <- dscore(1, items, age, full = TRUE, lexicon = "gcdg", qp = qp)
fp1 <- data.frame(qp = qp, fp = fp[[1]]$posterior, distribution = rep("posterior", length(qp)))
prior <- data.frame(qp = qp, fp = fp[[1]]$start, distribution = rep("prior", length(qp)))

dataprior <- rbind(fp1, prior)
dataprior$distribution <- as.factor(dataprior$distribution)

p1 <- ggplot(dataprior, aes(qp, fp, group = distribution)) +
      geom_path(aes(col = distribution)) +
      coord_cartesian(xlim=c(-10, 80)) +
      scale_y_continuous(name = "Density") +
      scale_x_discrete(name = "D-score")
p1
```

Now, let's add an item on which the scores for the two example
children differ, e.g. item *n35*. Item *n35* is also a gross motor
item, namely 'walks along'. So both children are able to crawl (with
their abdomen of the floor), but only *child 1* is able to walk along.
We can use the previously calculated posterior distribution as prior
when we add item *n35*.

Let's start with *child 1*, who is able to walk along. Using Bayes
rule, the new posterior distribution can be calculated. In figure
\@ref(fig:fig2itemC1) the prior, in this case the posterior after
adding item *n34* to the age-dependent prior, and the newly calculated
posterior distribution are visualized. Again, the posterior and prior
are centred around the same point, while the posterior has a bit more
precision.

```{r, fig2itemC1, echo=FALSE, fig.cap="\\label{fig:fig2itemC1} Prior and posterior in the 2nd step of the iterative process for child 1"}
items <- c("n34", "n35")
age <- rep(1.25, length(items))
  
#full posterior 2e item
qp <- -10:80
fp2 <- dscore(c(1,1), items, age, full = TRUE, lexicon = "gcdg", qp = qp)
fp1 <- data.frame(qp = qp, fp = fp[[1]]$posterior, distribution = rep("1st posterior", length(qp)))
fp2 <- data.frame(qp = qp, fp = fp2[[1]]$posterior, distribution = rep("2nd posterior", length(qp)))

datafp <- rbind(fp1, fp2)
datafp$distribution <- as.factor(datafp$distribution)

p2 <- ggplot(datafp, aes(qp, fp, group = distribution)) +
      geom_path(aes(col = distribution)) +
      coord_cartesian(xlim=c(-10, 80)) +
      scale_y_continuous(name = "Density") +
      scale_x_discrete(name = "D-score")
p2


```

Now we can do the same thing for *child 2*, who is able to crawl with
his/her abdomen off the floor (*n34*) but not to walk along (*n35*).
Again, we can use the posterior after adding information from *n34* as
the prior and use the information from *n35* to calculate the
posterior. In figure \@ref(fig:fig2itemC2) the prior and posterior are
given. Now the prior and posterior are different, the posterior is on
the left of the prior and has an high precision in comparison to the
prior. Hence, being unable to walk along at the age of 15 months, like
*child 2*, is more indicative for the development of a child of 15
months than being able to walk along like *child 1*.

```{r, fig2itemC2, echo=FALSE, fig.cap="\\label{fig:fig2itemC2} Prior and posterior in the 2nd step of the iterative process for child 2"}
items <- c("n34", "n35")
age <- rep(1.25, length(items))
  
#full posterior 2e item FAIL
qp <- -10:80
fp2 <- dscore(c(1,0), items, age, full = TRUE, lexicon = "gcdg", qp = qp)
fp1 <- data.frame(qp = qp, fp = fp[[1]]$posterior, distribution = rep("1st posterior", length(qp)))
fp2 <- data.frame(qp = qp, fp = fp2[[1]]$posterior, distribution = rep("2nd posterior", length(qp)))

datafp2 <- rbind(fp1, fp2)
datafp2$distribution <- as.factor(datafp2$distribution)

p3 <- ggplot(datafp2, aes(qp, fp, group = distribution)) +
      geom_path(aes(col = distribution)) +
      coord_cartesian(xlim=c(-10, 80)) +
      scale_y_continuous(name = "Density") +
      scale_x_discrete(name = "D-score")
p3

```


The next four steps in the iterative procedure, information from the
other four items (*n32*, *n33*, *v38* and *n37*) can be added in a
similar fashion. Figure \@ref(fig:fig6itemC1C2) shows the final
posterior distributions for *child 1* and *child 2* and the prior
distribution for the age of 15 months. The D-score can now be
determined by looking at the mean of the posterior distributions.
Hence, the D-score for *child 1* and *child 2* are equal to 55.75 and
47.76, respectively.

```{r, fig6itemC1C2, echo=FALSE, fig.cap="\\label{fig:fig6itemC1C2} Prior and final posterior for child 1 and 2"}
items <- c("n32", "n33", "v38", "n37","n34", "n35")
age <- rep(1.25, length(items))
  
#full posterior for both children and prior
qp <- -10:80
fp6_1 <- dscore(c(1,1,1,1,1,1), items, age, full = TRUE, lexicon = "gcdg", qp = qp)
fp6_2 <- dscore(c(1,0,1,1,1,0), items, age, full = TRUE, lexicon = "gcdg", qp = qp)
prior <- data.frame(qp = qp, fp = fp[[1]]$start, distribution = rep("prior", length(qp)))
fp6_1 <- data.frame(qp = qp, fp = fp6_1[[1]]$posterior, distribution = rep("final posterior child 1", length(qp)))
fp6_2 <- data.frame(qp = qp, fp = fp6_2[[1]]$posterior, distribution = rep("final posterior child 2", length(qp)))

datafp6 <- rbind(prior, fp6_1, fp6_2)
datafp6$distribution <- as.factor(datafp6$distribution)

p6 <- ggplot(datafp6, aes(qp, fp, group = distribution)) +
      geom_path(aes(col = distribution)) +
      coord_cartesian(xlim=c(-10, 80)) +
      scale_y_continuous(name = "Density") +
      scale_x_discrete(name = "D-score")
p6

```

```{r, echo=FALSE, message=FALSE}
#Dscore child1
dscore(c(1,1,1,1,1,1), items, age, full = FALSE, lexicon = "gcdg") #55.75
#Dscore child2
dscore(c(1,0,1,1,1,0), items, age, full = FALSE, lexicon = "gcdg") #47.76
```

*nog toe te voegen in de tekst*

*quadratic points*

*Note: met deze methode is het mogelijk om scores voor extremen te berekenen, andere methodes (zoals de worms estimator) kunnen dit niet altijd.*

*Prior die gekozen was: plafond prior (wanneer je alle items kan)*


We already explained how the D-score suffers from a ceiling and bottom
at the start and end of the scale. However, this can also happen
within the D-score scale: some values of the D-score will never occur
no matter the combination of scores on the milestones. This will
mainly occur when calibrated milestones are used to calculate the
D-score. For example, the Dutch Developmental Instrument (DDI), which
is used as a screening instrument, is calibrated: at each wave 90\% of
the children should be able to get a positive score on the milestones.
Which allows for detection of children with a developmental delay.
