# Validity {#ch:validity}

> Author: Iris Eekhout

Validity is a generic term that refers to the question how well an
instrument measures what it claims to measure. There are many various
aspect of validity. This chapter briefly reviews the main types of
validity:

* Internal validity (\@ref(sec:internal))
* External validity (\@ref(sec:external))


```{r init07, include=FALSE}
library(ddata)
library(gseddata)
library(dplyr)
library(tidyr)
```

```{r smoccmodel7, dependson = "smoccmodel", cache = FALSE}
data <- knitr::load_cache("smoccmodel", "data")
items <- knitr::load_cache("smoccmodel", "items")
model <- knitr::load_cache("smoccmodel", "model")
```

## Internal validity {#sec:internal}

### Content validity

Content validity is the extent to which the $D$-score represents all
facets of development. In contrast to "face validity", which assesses
whether the test appears valid to respondents, content validity is
about what is actually measured. 

One important form of content validity is that we wish to make sure
that the measurement scale represents the various developmental
domains in a fair way. In the simplest case, we can assign each
milestone uniquely to one domain, and evaluate coverage by splitting
the cumulative item information over domains.

```{r plotcontent, fig.height=3, fig.show='hold', message=FALSE, warning=FALSE, fig.cap = '(ref:plotcontent)'}
info_data <-
  dinstrument::info_d_item(
    itembank = model$itembank,
    delta = "tau",
    long = TRUE,
    beta_range = 0:80
  )
dom_data <- info_data %>% 
  mutate(domain = gseddata::decompose_itemnames(lex_gsed)[, "domain"],
         domain = plyr::revalue(domain, c("cm" = "Communication", 
                                          "fm" = "Fine motor",
                                          "gm" = "Gross motor")))
ggplot(data = dom_data, aes(x = ability, y = info, fill = domain)) +
  geom_bar(stat = "identity") +
  ylab("Item information") +
  xlab("Ability (D-score)") + 
  labs(fill = "DDI domain") +
  theme(legend.position = c(0.15, 0.75))
```

(ref:plotcontent) Cumulative item information by DDI domain.

Figure \@ref(fig:plotcontent) shows the coverage of the three domains
of the DDI at various levels of the $D$-score. The three domains of
the DDI are represented at most ability levels. The DDI contains no
communication milestones between 20 $D$ and 30 $D$, so at these levels
the DDI measures primarily motor performance.

### Construct validity

Construct validity is the extent to which the $D$-score behaves like
the theory says the construct should behave. For example, we expect
that child development advances with by age. Figure \@ref(fig:smoccda)
provides convincing evidence that the $D$-score increases fastest in
the first six months, and keeps rising at a slower rate as children
age. This is consistent with theories in growth and child development.

```{r plotdomainfit, fig.height=6, fig.show='hold', message=FALSE, warning=FALSE, fig.cap = '(ref:plotdomainfit)'}

fitdomplot <- model$item_fit %>%
  mutate(lex_gsed = item,
         domain = gseddata::decompose_itemnames(item)[, "domain"],
         domain = plyr::revalue(domain, c("cm" = "Communication", 
                                          "fm" = "Fine motor",
                                          "gm" = "Gross motor"))) %>%
  left_join(model$itembank[,c("lex_gsed", "tau")], by = "lex_gsed") %>%
  gather(key = "type", value = "fit", infit, outfit)


ggplot(data = fitdomplot, aes( x = tau, y = fit, color = domain)) +
  geom_point() +
  facet_grid( type ~.)+
  xlab("D-score") +
  ylab("Item fit")

```

(ref:plotdomainfit) Item fit by $D$-score for the DDI domains.

In Chapter \@ref(ch:newmodel) we assumed that child development can be
measured by one latent variable. Figure \@ref(fig:plotdomainfit)
provides one way to evaluate the validity of this assumption. The
figure plots the item fit for each milestone colored by domain. Items
from different domains fit equally well, so there is no evidence that
the $D$-score is tilted towards a particular domain. Put in more
technical terms, the DDI domains do not explain differences in the
item fit residuals of the model.

## External validity {#sec:external}

### Discriminatory validity

---
# >> achtergrondvariabelen gebruiken zoals opleidingsniveau ? Disclamer toevoegen dat de informatie in de data summier is, maar dat we nog wel meer zouden willen toevoegen. 
---

Discriminatory validity indicates the extent to which the $D$-score
can distinguish children with non-normal development from children
that are developing normally. We may evaluate this by identifying
children with lagging development, for example indicated by reflex
and/or tonus problems, and study whether the $D$-score is able to
discriminate those children from the general population. Section
\@ref(sec:togo) presents some examples.

```{r plotdiscr, eval=FALSE, fig.height=3, fig.show='hold', message=FALSE, warning=FALSE}
plotdat <- left_join(model$dscore, data$visit) %>%
  mutate(agemos = agedays /365.25 * 12,
         edumocat = factor(edumocat, labels = c("none", "primary", "secondary", "above secondary")),
         residence = factor(residence, labels = c("semi-urban", "urban", "metropolitan")))

ggplot(plotdat, aes(x = agemos, y = b, group = edumocat, colour = edumocat))+
  geom_point(cex = 0.8) +
  geom_smooth() +
  xlab("Age (months)")+
  ylab("D-score")


ggplot(plotdat, aes(x = agemos, y = b, group = residence, colour = residence))+
  geom_point(cex = 0.6)+
  xlab("Age (months)")+
  ylab("D-score")
```


### Convergent and divergent validity

Convergent validity is the extent to which the $D$-score relates to
similar constructs. It can be measured by the correlation with other
instrument that measure development. Examples include the total score
for Bayley-III or the Denver.

For a good convergent validity the correlation with other construct
should be 0.6, or higher. Unfortunately, at present only limited data
is for the DDI, so we cannot assess convergent validity for the
$D$-score at this point.

Divergent validity is the extent to the $D$-score is uncorrelated 
with measures that are different construct. 

```{r plotconcurrent, eval=TRUE, fig.height=6, fig.show='hold', message=FALSE, warning=FALSE, fig.cap = '(ref:plotconcurrent)'}
cor1dat <- left_join(model$dscore, data$visit) %>% 
  mutate(daz = dscore::daz(b, 
                           x = .data$agedays, x.unit = "day"))
cor1 <- round(cor(cor1dat[,c("b", "htcm")], use = "complete.obs")[1,2],2)
cor2 <- round(cor(cor1dat[,c("daz", "haz")], use = "complete.obs")[1,2],2)

dmetric::plot_d_adm(data = data, model = model, metric = "dscore", 
                    adm_list = "htcm", show_legend = FALSE,
                    show_cor = FALSE)[[1]] +
  annotate("text", x = 60, y = 58, hjust = 0, label = paste("r(D, Height) =", cor1)) +
  annotate("text", x = 60, y = 54, hjust = 0, label = paste("r(DAZ, HAZ) =", cor2)) +
  ylab("Height (cm)")

#plot_d_adm(data = data, model = model, metric = "daz", adm_list = "haz", show_legend = FALSE, show_cor = FALSE)[[1]]+
#  annotate("text", x = 2, y = -5.5, hjust = 0, label = paste("correlation =", cor2))+
#  ylab("z-score for height")+
#  xlab("z-score for D")
```

(ref:plotconcurrent) Relation between body height and the $D$-score in the 
SMOCC data.

Figure \@ref(fig:plotconcurrent) shows both convergent and divergent validity at work. 
The figure shows that, as expected, there is a strong and almost linear relation between body height and the $D$-score. However, after correction for child's age, the relation between height and $D$-score almost disappears. This indicates that, given age, growth and development are quite different concepts.

We can also evaluate the strength of the relation between the
$D$-score and measures that are currently used as a proxy for child
development, such as stunted height growth (see section
\@ref(sec:stunting)). The low correlation between DAZ and HAZ suggests
that stunting is a bad proxy for child development.

### Predictive validity

Predictive validity refers to the degree to which the $D$-score predicts
score on a criterion that is measured late in time. For the $D$-score
we have measures for IQ at the age of 5 as a possible criterion.

@vlasblom2019 found strong evidence that individual milestones of the
  DDI measured during the first years of life are predictive of later
  attendance of special education at ages 5--10 years. It is to be
  expected that the $D$-score, which is built upon these items, will
  also predict limited intellectual functioning.
