---
output: pdf_document
---
```{r init11, include=FALSE}
source(file = file.path("/Users/buurensv/Websites/dbook/dbook1/R", "before_chapter_script.R"), 
       local = TRUE)
```

# Next steps {#ch:nextsteps}

> Author: Stef van Buuren

This booklet provides a quick overview of the relevance, concepts and
techniques behind the $D$-score. While the results obtained thus far
are encouraging, some questions will certainly remain when the method
is put to practice.

A rough selection of such questions includes:

* What is the added value of the $D$-score in practice? 
* Does the $D$-score extend to higher ages?
* Is the assumption of uni-dimensionality reasonable for other ages and populations?
* Can the $D$-score be calculated from instruments other than the DDI?
* Is it reasonable to assume that milestone difficulty is identical in other populations?
* Does the method apply to caregiver-reported milestones?
* Would a dedicated $D$-score instrument be more efficient? 
* How many milestones are "enough"?
* Can the same scale be used for measurement at individual, group and population levels?
* Can the $D$-score detect delayed development?
* Would the $D$-score help to target early interventions?

This chapter briefly reviews some of these issues.

## Usefulness of $D$-score for monitoring child health

The $D$-score is a new approach to measure child development. The
$D$-score is a scale for quantifying generic child development by a
single number. The $D$-score fits the Rasch model. The measurements
can therefore be interpreted as scores on an interval scale, a
requirement for answering questions like:

* What is the difference in development over time for the same child,
group or population?

* What is the difference in development between different children,
groups or populations of the same age?

* How does child development compare to a norm?

The $D$-score is not tied to a particular instrument. Any instrument
that fits the model underlying the $D$-score can be used to measure
the child's $D$-score.

The precision of the measurement depends on the number of milestones
used, as well as on the match between the difficulty of the milestones
and the ability level of the measured child. The measurement
instrument can thus be tailored to the question at hand.

## $D$-chart, a growth chart for child development

The field of child growth and development roughly divides into two
areas:

* The subfield *child growth* (or *auxology*) emphasizes body measures
like height, weight, body mass index, and so on. It is a rigorous
quantitative science with intimate ties to statistics since the days
of Quetelet and Galton.

* The subfield *child development* is more recent and builds upon a
wide-ranging set of domain-specific instruments for measuring motor,
language, cognitive and behavioral states. 

The *growth chart* is a widely used tool to monitor physical growth.
The $D$-score can be used in a similar way to create the *$D$-chart*.

```{r smoccmodel10, dependson = "smoccmodel", cache = FALSE}
data <- knitr::load_cache("smoccmodel", "data")
items <- knitr::load_cache("smoccmodel", "items")
model <- knitr::load_cache("smoccmodel", "model")
```

```{r dchart, warning = FALSE, fig.cap = '(ref:dchart)', fig.height = 6}
# set.seed(81104)
set.seed(3)
kids <- sample(unique(model$dscore$subjid), 5)
data <- model$dscore %>% 
  filter(subjid %in% kids) %>% 
  mutate(subjid = as.factor(subjid), 
         agemos = agedays / 365.25 * 12)

reference <- dplyr::select(dscore::Dreference, month, SDM2:SDP2) %>%
  filter(month <= 30) %>%
  tidyr::gather(key = centile, value = d,-month)
polygon <- data.frame(
  x = c(reference$month[reference$centile == "SDM2"],
        rev(reference$month[reference$centile == "SDP2"])),
  y = c(reference$d[reference$centile == "SDM2"],
        rev(reference$d[reference$centile == "SDP2"])))

g1 <- ggplot(data, aes_string(x = "agemos", y = "b", colour = "subjid")) +
  scale_x_continuous("Age (in months)",
                     limits = c(0, 30),
                     breaks = seq(0, 30, 6)) +
  scale_y_continuous("D-score",
                     limits = c(0, 75),
                     breaks = seq(0, 70, 10)) +
  geom_polygon(mapping = aes(x = x, y = y), 
               data = polygon, col = "transparent", fill = "grey", alpha = 0.1) + 
  geom_line(mapping = aes(x = month, y = d, group = centile), 
            data = reference, colour = "grey", alpha = 0.3) +
  geom_point() +
  geom_line() + 
  theme(legend.position = c(0.8, 0.2))
g1
```

(ref:dchart) $D$-chart with five children from the SMOCC study.

Figure \@ref(fig:dchart) shows developmental trajectories of five
randomly chosen children from the SMOCC study. Note that the
trajectory can be created even though the milestones are different at
different ages. The trajectory may help to track development of a
child on a periodic basis.

The $D$-chart shows that it is straightforward to apply quantitative
techniques from child growth to child development. Our hope is that
$D$-score aids in bridging the somewhat disparate subfields of child
growth and child development.

## Opportunities for early intervention

@black2017 estimated that about 250 million children worldwide fail to reach
their developmental potential. Developmental delays become evident in
the first year, and worsen during early childhood. The burden of
children not reaching their developmental potential is high.

Interventions aimed at improving child development should be delivered
at the sensitive periods for maximum effect. Programs are to be
comprehensive, incorporating a combination of health, nutrition,
security and safety, responsive care-giving and early learning. See
@engle2011 and @grantham2014 for recent overviews.

Both estimation of intervention effects as well as comparison across
different treatments is hampered by the absence of universal measure
for child development. The lack of a common yardstick may have adverse
impact on proper policy and priority setting.

Since the $D$-score is not tied to a particular instrument, our hope
is that it may help in improving decision making, ultimately lowering
the number of children not reaching their developmental potential.

## $D$-score for international settings

Chapter \@ref(ch:threestudies) compared $D$-scores between three study
samples. The analysis was simplified by the fact that all three
studies used the same instrument (the DDI, in Togo translated to
French) to measure child development. 

It is difficult to compare levels of child development worldwide.
Existing estimates on children not reaching their developmental
potential rely on proxies, such as stunting and poverty. While these
proxies have been found to correlate with child development, they are
only weak indicators of actual child performance. Arguably, the
performance of a child on a set of well-chosen milestones is more
informative for his or her future health and productivity than body
height or parental income.

One way to improve the quality on child development estimates is to
*use the same instrument worldwide*. While simple in theory, such a
pre-harmonization approach is difficult in practice, and may require
huge and continued investments by a central agency. Moreover, it does
not address historic data, so it will be difficult to see secular
trends.

The $D$-score opens up an interesting alternative: *agree on the
scale*, and leave some liberty to the data-collector in the exact
choice of the instrument. This may be a better use of the
data-collector's expertise of the local population, and may be more 
successful in keeping up with innovations in measurement.

Booklet II will address some of the conceptual and technical issues
that arise when we attempt to apply the $D$-score to other
populations.

## $D$-score from existing instruments

There is a huge base of historic child developmental data using, of
course, existing instruments. The problem is that each instrument
defines its own summaries, and these summaries cannot be compared
across instruments.

Different instrument have different domains, different age forms,
different stopping rules, different age norms, and so on. Yet, the
milestones in these instruments are often very similar. Most
instruments collect data on milestones like:

* Can the child stack two blocks?
* Can the child roll over?
* Can the child draw a cross?
* Can the child stand?
* Can the child say "baba"?

With the $D$-score methodology in hand, we are ready to exploit the
overlap in common milestones from different instrument. Common items
can act as *bridges* between instruments, so - with the appropriate
item-level data - we may attempt calculating $D$-scores from other
instruments as well.

Booklet II will explore this possibility in more detail. 

## Creating new instruments for $D$-score

Extending the $D$-score to other instruments has the side-effect of
extending the itembank with good items. Moreover, as more and more
data feed into the itembank, the estimates of existing may become more
precise. 

The enlarged and improved itembank then may act as the basic resource
for creating instruments for particular settings. For example, if the
interest is on finding the most advanced children, we may construct a
difficult test that will separate the good and the best. Alternatively,
we can use the item bank to create and administer *computerized
adaptive tests* [@wainer2000, @jacobusse2007], a sequential method that select the
next milestone based on the previous test outcome.

Booklet III will explore the conceptual and technical challenges, and
propose an integrated approach to support instrument construction and 
validation.
